A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-24 tex/BiSig.tex
[ fix ] The premise of the annotation rule
[ update ] Simplified notations

\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage[all,cmtip]{xy}
\usepackage{bussproofs}
\EnableBpAbbreviations

@@

\input{macros}

@@

@@

\maketitle
\section{Introduction}

@@

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}
\subsection{Type Checking via Bidirectional Typing}

\section{A Definition of Simple Type Theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple) type signature} is a set $I$ with an \emph{arity} $\mathit{ar}_i : \mathbb{N}$ associated to each $i : I$.
  An element $i : I$ is meant to be the \emph{$i$-th operation} $\tyop{i}$ and $\mathit{ar}(i)$ the \emph{number} of arguments of $\tyop{i}$.
\end{definition}
\LTnote{Define $\mathsf{Type}$}
\begin{figure}
  \begin{subfigure}{.5\textwidth}
    \begin{mathpar}
      \boxed{\Theta\vdash A} \\
      \inferrule{\Theta \ni X_i}{\Theta \vdash X_i} \and
      \inferrule{\Theta \vdash A_1 \\ \cdots \\ \Theta \vdash A_n}{\Theta \vdash \tyop{i}(A_1, \ldots, A_n)}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}{.4\textwidth}
    \begin{mathpar}
      \boxed{\Theta \vdash \Gamma} \\
      \inferrule{ }{\Theta \vdash \emptyset}\and
      \inferrule{\Theta \vdash A \\ \Theta\vdash \Gamma}{\Theta \vdash \Gamma, A}
    \end{mathpar}
  \end{subfigure}
  \caption{Simple types and contexts}
\end{figure}

\begin{example}[STLC]

@@

\subsection{Binding Signatures and Typed Terms}
\begin{definition}
  For a type signature, a \emph{term signature} consists of a set $J$ meant to be the set of term constructs $\op{j}$ with a triple $\left(\Xi_j, A_{0, j}, \left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]\right)$ for each $j:J$ where
  \begin{enumerate}
    \item $A_{0, j} : \mathsf{Type}(\Xi_j)$ is the type of its target and
    \item $\left[\left(\Delta_{1, j}, A_{1, j}\right), \ldots, \left(\Delta_{n_j, j}, A_{n_j, j}\right) \right]$ is a finite sequence of pairs of a context $\Delta_{k, j} : \mathsf{Cxt}(\Xi_j)$ for types of its binding variables and its type $A_{k, j} : \mathsf{Type}(\Xi_j)$.
  \end{enumerate}
\end{definition}
\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash t : A} \\
    \inferrule{\Gamma \ni (x : A)}{\Theta \mid \Gamma \vdash \term{x} : A}\;(\text{var})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\Delta_{1,j}}{\rho} \vdash \term{t_1} : \simsub{A_{1,j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n_j, j}}{\rho} \vdash \term{t_{n}} : \simsub{A_{n, j}}{\rho}} {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(t_1, \ldots, t_{n_j})} : \simsub{A_{0,j}}{\rho}}\;(\text{op})
    \quad \text{where $\rho \colon \Xi_j \to \Theta$}
  \end{mathpar}

  \caption{Typing rules}
\end{figure}

\begin{example}[STLC]

@@

\begin{definition}

\end{definition}

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \term{t} \Rightarrow A} \quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash \term{t} \Leftarrow A} \\
    \inferrule{\Gamma \ni (\term{x} : A)}{\Theta \mid \Gamma \vdash \term{x} \mathrel{\color{red}\Rightarrow} A}\;(\text{var}{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash (\term{t} : B) \Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash \term{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \term{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\Theta \mid \Gamma, \simsub{\vec{x}_{1,j} : \Delta_{1, j}}{\rho} \vdash \term{t_{1, j}} \mathrel{\color{red}\leftrightarrows_{1, j}} \simsub{A_{1, j}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\vec{x}_{n_j} : \Delta_{n_j, j}}{\rho} \vdash \term{t_{n_j, j}} \mathrel{\color{red}\leftrightarrows_{n_j}} \simsub{A_{n_j ,j}}{\rho}}
    {\Theta \mid \Gamma \vdash \term{\op{j, \rho}(\vec{x}_1 .\, t_1, \ldots, \vec{x}_{n_j}.\, t_{n_j})} \mathrel{\color{red}\leftrightarrows_{0, j}} \simsub{A_{0, j}}{\rho}} \; (\text{op}{\leftrightarrows})
    \quad \text{where $\rho \colon \Xi_i \to \Theta$}
  \end{mathpar}
  \caption{Bidirectional typing rules}
\end{figure}

\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}
\subsubsection{Intrinsically Typed Terms}

\subsubsection{Raw Terms}

\begin{figure}
  \caption{Bidirectional raw terms}
\end{figure}
\subsubsection{Extrinsically Typed Terms}

@@

\section{Bidirectional Type Synthesis and Checking}
\subsection{Substitution Properties}
\subsection{Generic First-Order Unification}
\cite{McBride2003,McBride2003a}
\subsection{Uniqueness of Synthesis}

>>>>

\usepackage{bm}
\usepackage{braket}
\usepackage{mathtools}
\usepackage{mathpartir}
\usepackage[all,cmtip]{xy}
\usepackage{bussproofs}
\EnableBpAbbreviations

@@

@@

\AtEndPreamble{%
\theoremstyle{acmdefinition}
\newtheorem{remark}[theorem]{Remark}}

\input{type-notation} %% for type theory notation copied from HoTT Book
\input{macros}

@@

\maketitle
\section{Introduction}\label{sec:intro}

@@

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

\subsection{Type Synthesis and Checking via Bidirectional Typing}
\subsection{Formalisation in \Agda}

\section{A Definition of Simple Type Theories}\label{sec:simple-type-theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.

  The judgements of \emph{$\Sigma$-types} and \emph{$\Sigma$-contexts} over a variable set $\Theta$ are defined inductively by \Cref{fig:simple-type,fig:simple-context} respectively.
We write $A : \Type_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} A$ and $\Gamma \colon \Cxt_{\Sigma}(\Theta)$ for $\Theta \vdash_{\Sigma} \Gamma$ to disambiguate $\Sigma$-types and $\Sigma$-contexts in the case of any confusion.
\end{definition}

\begin{figure}
  \begin{minipage}[b]{.55\textwidth}
    \centering
    \small
    \begin{mathpar}
      \boxed{\Theta\vdash_{\Sigma} A} \\
      \inferrule{\Theta \ni X_i}{\Theta \vdash_\Sigma X_i} \and
      \inferrule{\Theta \vdash_{\Sigma} A_1 \\ \cdots \\ \Theta \vdash_{\Sigma} A_n}{\Theta \vdash_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}
    \end{mathpar}
    where $n = \arity_i$
    \caption{Type formation}
    \label{fig:simple-type}
  \end{minipage}
  \begin{minipage}[b]{.4\textwidth}
    \centering
    \small
    \begin{mathpar}
      \boxed{\Theta \vdash_{\Sigma} \Gamma} \\
      \inferrule{ }{\Theta \vdash_{\Sigma} \cdot }\and
      \inferrule{\Theta \vdash_{\Sigma} A \\ \Theta\vdash_{\Sigma} \Gamma}{\Theta \vdash_{\Sigma} \Gamma, A}
    \end{mathpar}
    \caption{Context formation}
  \label{fig:simple-context}
  \end{minipage}
\end{figure}

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically has a binary implication type $\bto$ and some base types so that the set of all types is at least inhabited.
  In our setting, the set of $\Sigma$-types over $\Theta$ may have variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is inhabited if $\Theta$ is.
  Hence, there is no need to introduce base types which are not introduced by any term construct.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus is thus given solely by an operation $\bto$ with $\arity(\bto) = 2$, i.e.\ a binary operation.
  For simply typed $\lambda$-calculus with binary products, we can augment
  $\Sigma_{\bto}$ by adding a binary operation $\btimes$ for the binary product type.
  We denote this augmented signature by $\Sigma_{\bto, \btimes}$.

@@

\subsection{Binding Signatures and Intrinsically Typed Terms}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\termOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\termOp_o$ and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\termOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

\begin{example} \label{ex:STLC-sig}
  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
  \end{align*}
    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
  \begin{align*}
    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
    && 
    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
  \end{align*}
\end{example}

\begin{example}\label{ex:STLCx-sig}
  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
to $\Omega_{\bto}$:
  \begin{align*}
    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
    && \fst \colon A, B \vdash A \btimes B \to A
    && \snd \colon A, B \vdash A \btimes B \to B
  \end{align*}
\end{example}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma_{\Sigma, \Omega} \vdash A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho : \Xi \to \Theta \\\Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
\end{figure}

\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
  \end{mathpar}

  \caption{Raw terms}
\end{figure}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, \leftrightarrows_1, A_1\right), \ldots, \left(\Delta_{n}, \leftrightarrows_n, A_{n}\right) \right], \leftrightarrows, A\right)
  \]
  where ${\leftrightarrows_i}$ and ${\leftrightarrows}$ are of type $\Mode$ indicating the modes of a construct and its arguments respectively.
  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ to indicate an operation $o$ with its arity. 
\end{definition}

\begin{example}
  With the type signature $\Sigma_{\bto}$, the term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  like those in \Cref{ex:STLC-sig} but augmented with the mode information.
\end{example}

\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}

\subsubsection{Intrinsically Typed Terms}

\subsubsection{Bidirectional Raw Terms}

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Rightarrow}
    \quad\text{and}\quad
    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Leftarrow}
  \end{mathpar}

  \caption{Bidirectional raw terms}
\end{figure}

\subsubsection{Extrinsically Typed Terms}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\rho\colon \Xi \to \Theta \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
  \end{mathpar}
  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

@@

\section{Bidirectional Type Synthesis and Checking}

\subsection{Substitution Properties}
\subsection{Generic First-Order Unification}
\cite{McBride2003,McBride2003a}
\begin{definition}
A \emph{decidable equality} for a type $X$ is a dependent function $\prod_{x, y : X} \left(\id{x}{y}\right) + \neg \left(\id{x}{y}\right)$.
\end{definition}

\subsection{Uniqueness of Synthesis}
